---
title: "car race deaths data"
author: "YOUR NAMES, NAMES, NAMES" 
date: "10/20/2020"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Just write a paragraph where you explain the purpose of this document. REVERSE ENGINEER THE STORY. [Out of control](https://www.latimes.com/projects/la-me-street-racing/).

## Load libraries and options

```{r}
# Get rid of scientific notation
options(scipen = 999)

# install.packages("sf")

# For cleaning data
library(janitor)
#To load the data from reporters
library(tidyverse)
library(sf)
library(lubridate)

```

## Load and clean data

DESCRIPTION OF WHAT THE DATA IS, WHERE YOU GOT IT, LINK IT, WHAT IT INCLUDES, INCLUDING A LIST OF WHAT EACH FIELD IS. 

* age | number | age of driver
Street Racing Columns
Age 
Case Number
DeathDateTime
Gender 
Names
Race
Victime ROles
Full Event Address
Geocoder address

fatalities
name
city 
population
geometry 



```{r}
# Comment on what you're loading
street_racing <- read_csv("input/street-racing-fatalities.csv")

# Comment on what you're loading
fatalities_racing <- read_csv("output/fatalities.csv")

# Comment on what you're loading
# Special form of geographic data called .shp
citypop <- st_read("input/hoods.shp")%>%
  as_tibble() %>%
  select(-geometry)

```

## Analysis

### Sentence 1

Deaths have occurred all across L.A. County, with the biggest concentration in the small city of Commerce. Eleven have died on the small, industrial city’s streets since 2000.

```{r}
# Comment explaining what is happening in a given codeblock
fatalities_truncated <- fatalities_racing%>%
  select(name, city, death_datetime, death_year)

fatalities_truncated
```

```{r}
# Comment explaining what is happening in a given codeblock
commerce_deaths <-fatalities_truncated%>%
  filter(name == "Commerce", death_year >= "2000")

commerce_deaths

```



```{r}
# Comment explaining what is happening in a given codeblock
death_by_city <- fatalities_truncated%>%
  group_by(name)%>%
  summarise(city_deaths = n())%>%
  arrange(desc(city_deaths))

death_by_city
```

```{r}
# This code joins the citypop table we created with the death by city frame and then creates a new column dividing the number of deaths per city by the population (multiplying by 10,000) and then filters out any cities with populations less than a thousand, generating the 4th finding: Deaths have occurred all across L.A. County, with the biggest concentration in the small city of Commerce. Eleven have died on the small, industrial city’s streets since 2000.

death_by_city_pop <- death_by_city%>%
 inner_join(citypop)%>%
  mutate(deaths_per_capita = (city_deaths/population*10000))%>%
  filter(population > 1000)%>%
  arrange(desc(deaths_per_capita))


death_by_city_pop

```
### Sentence 2

The deadliest year on record was 2007, with 18 fatal crashes. After a period with relatively few recorded deaths, the count grew in recent years, with 15 fatalities in 2015, 11 in 2016 and 12 deaths in 2017, the Times analysis shows.

```{r}
deaths_by_year <- street_racing%>%
  clean_names()%>%
  mutate(year = year(death_date_time))%>%
  select(year, case_num, name_first, name_last, victim_role)
 
deaths_by_year


```
```{r}

# This codeblock reverse-engineers a portion of the following finding: "The deadliest year on record was 2007, with 18 fatal crashes. After a period with relatively few recorded deaths, the count grew in recent years, with 15 fatalities in 2015, 11 in 2016 and 12 deaths in 2017, the Times analysis shows.

#This codeblock shows that "the deadliest year on record was 2007, with 18 fatal crashes" by: creating a new object called deaths_grouped and then groups the deaths_by_year table by year, and then counts the number of deaths in each year, and then arranges the total_deaths column from highest to lowest. 
deaths_grouped<- deaths_by_year%>%
  group_by(year)%>%
  summarise(total_deaths = n())%>%
  arrange(desc(total_deaths))

deaths_grouped

```

```{r}

# This codeblock reverse-engineers a portion of the following finding: "The deadliest year on record was 2007, with 18 fatal crashes. After a period with relatively few recorded deaths, the count grew in recent years, with 15 fatalities in 2015, 11 in 2016 and 12 deaths in 2017, the Times analysis shows.

#This codeblock shows that "after a period with relatively few recorded deaths, the count grew in recent years, with 15 fatalitiesin 2015, 11 in 2016 and 12 deaths in 2017," by: creating a new object called deaths_grouped_again and then grouping the deaths_by_year table by year, and then counting the number of deaths in each year, and displaying a table that shows that from 2009-2013, streetracing deaths were in the single digits until 2014, when they hit 11 and then 2015, with 15 deaths. 

deaths_grouped_again<- deaths_by_year%>%
  group_by(year)%>%
  summarise(total_deaths = n())

deaths_grouped_again
```
### Sentence 3

The dead were overwhelmingly male and young: More than half were 21 or younger, including two children, ages 4 and 8, killed along with their mother. Slightly less than half of the victims— 47% — were behind the wheel. The rest were either passengers in the speeding vehicles, spectators or people simply walking on a sidewalk or driving home.

```{r}
#Creates a new object called 'deaths_by_year_expanded' which grabs data from "streetracing"
#AND THEN cleans up names in object 
#AND THEN creates a new column called year which grabs data from column 'year'
#AND THEN selects columns : 'year', 'gender', 'age', 'case_num', 'name_first', 'name_last', 'victim_role'
deaths_by_year_expanded <- street_racing%>%
  clean_names()%>%
  mutate(year = year(death_date_time))%>%
  select(year, gender, age, case_num, name_first, name_last, victim_role)
  
deaths_by_year_expanded
```
```{r}

#this codeblock reproduces the finding that 47% of the victims were behind the wheel by grouping the victims by their role and then summarising the number of victims in each role - 85 of the 179 total victims were suspected racers (behind the wheel), which comes out to 47.4%, or 47% rounded. 

behind_wheel <- deaths_by_year_expanded%>%
  group_by(victim_role) %>%
  summarise(deaths = n()) %>%
  mutate(victim_class = case_when(
    victim_role == "Suspected Racer" ~ "suspected_racer",
    TRUE ~ "not_suspected_racer"
  )) %>%
  group_by(victim_class) %>%
  summarise(deaths = sum(deaths)) %>%
  pivot_wider(names_from = victim_class, values_from = deaths) %>%
  mutate(total_deaths = not_suspected_racer+suspected_racer) %>%
  mutate(pct_suspected_racer = suspected_racer/total_deaths*100)

behind_wheel
```
```{r}
#Create an object called "male_and_young" that grabs data from object "deaths_by_year_expanded"
#AND THEN create a column using feature 'mutate()' that will be called "age_group"
#AND THEN data will display 'young' for anyone that is younger than 27, and 'old' for anyone that is above 27
#AND THEN the information will be grouped together based off of gender and age group. 
#AND THEN creates a new column 'n' that will display the data will be counted together to show the overall number for both old and young/ male and female
#AND THEN arranges data in column 'n' from highest to lowest
male_and_young <- deaths_by_year_expanded %>%
  mutate(age_group = case_when(
    age <= 27 ~ "young",
    TRUE ~ "old"
  )) %>%
  group_by(gender, age_group) %>%
  count() %>%
  arrange(desc(n))
  
male_and_young
```
```{r}
male_and_young <- deaths_by_year_expanded %>%
  mutate(age_group = case_when(
    age <= 21 ~ "young",
    TRUE ~ "old"
  )) %>%
  group_by(gender, age_group) %>%
  count() %>%
  arrange(desc(n))
  
male_and_young

 
```

```{r}
#Clean name for data frame "streetracing"
#AND THEN arranges the data by age
#AND THEN filters information in death_date_time column for a specific date and time of "2007-10-08 17:50:00"
#AND THEN creates a new object called "four_year_old"
#AND THEN filter by age to show for ages 4 and up
#AND THEN select the column death_date_time
#AND THEN create a new object called "tragic_accident" that calls on information from "streeracing"
#AND THEN inner join object "tragic_accident" with object "four_year_old"
street_racing %>%
  clean_names() %>%
  arrange(age) 

street_racing %>%
  clean_names() %>%
  filter(death_date_time == "2007-10-08 17:50:00")



four_year_old <- street_racing %>%
  clean_names() %>%
  filter(age == 4) %>%
  select(death_date_time)

tragic_accident <- street_racing %>%
  clean_names() %>%
  inner_join(four_year_old)
  
```

